External priors for the next generation of CMB experiments.

### 2016-02-22 Notes/response.txt

Major comments:

- In my opinion the authors are over-discussing well known
degeneracies between cosmological parameters adding no new
information. They formulate the key question in the second-to-last
paragraph of the introduction but there is no answer at the end of the
reading. While acknowledging the importance of having priors on some
parameters to better constrain neutrinos and dark energy, I think the
authors should explain or highlight how those priors will be obtained,
i.e., which external experiments are needed and how they will help CMB
Stage-IV. Moreover, additional datasets like galaxies surveys have
already been proved to be useful by a number of papers in the
literature, the authors should highlight what is new in their results.
Following on these concerns, it is not clear to me if the authors have
an idea on how to get those improvements, for example how to get a
better prior on n_s with non-CMB data. I find the discussion simply
speculative. This impacts in particular the abstract, the results and
the conclusion sections.

The key issue raised here seems to be: How best to address the question: What external information can be used to improve projected constraints by breaking degeneracies? We think the referee is saying there are 2 parts to this answer: (i) quantify how much the added information on other parameters breaks degeneracies and (ii) give a sense of how we could get that information. The referee points out that we did not do a good job explaining the possibilities for (ii) and that for (i) at least qualitatively other papers have discussed degeneracies. We think this paper does advance the discussion of how degeneracies impact future constraints. The plots in Figs. 1-4 are a new way of answering this question and we think quite useful. We thought it important to explain the degeneracies in the text for completeness, but we agree with the referee that these explanations are not new. It is the quantitative information in the plots that is new.

The second issue is explaining how to get this information. In the case of mnu, we have added text about the importance of getting \tau from either 21 cm reionization measurements or low-l cmb polarization experiments. For N_eff, we point out that high-z surveys obtain more Fourier modes in the linear regime, and therefore the range of modes [k_min,k_max] that are measured precisely can be widened. It is this widening that will improve the constraints on n_s [and therefore N_eff]. We now emphasize that z>1 surveys are crucial for this. H_0 is clearer: there are already a number of efforts underway to improve direct measurements of H_0, so the plots here are useful in that they lend further importance to those measurements.

@@

- The baseline dataset in all the results is using CMB+BAO which means
already CMB + external information. This is confusing, we are not
improving on CMB then but rather on a combination of CMB and some
galaxy data.

The referee is right; we were not clear that CMB+BAO includes current BAO constraints, from e.g., BOSS.
It doesn't make sense to us to exclude things we currently know, so we think it important to include these constraints as our baseline. For example that will strongly bias the forecasted performance of CMB S4 when considered without future BAO experiments like DESI. We’ve added a line to make clear why we are including current BAO.

### 2016-02-22 Notes/CMBS4_priors_paper.tex
SD semi-final response

We also include information from current Baryon Acoustic Oscillation experiments and projected constraints from low-$l$ Planck polarization. 
For current BAO data (labeled BAO15), we follow \cite{allison:2015} and include 6dFGRS \cite{beutler:2011}, SDSS MGS \cite{ross:2015} together with LOW-Z and C-MASS \cite{anderson:2014}.
The noise assumed for Planck corresponds to the ``Planck-pol'' specifications of \cite{allison:2015}, where noise levels were approximated by scaling the current sensitivities according to the Planck Blue Book.

@@

%==========
\begin{figure}[htbp]
\begin{center}
\includegraphics{{prior_omnuh2_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the Hubble parameter, H$_{0}$. The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
\label{fig:prior_omeganuh2-h}
\end{center}
\end{figure}
%==========

%==========
\begin{figure}[htbp]
\begin{center}
\includegraphics{{prior_omnuh2_notau_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on $\Omega_{c} h^{2}$. The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
\label{fig:prior_omeganuh2-omch2}
\end{center}
\end{figure}
%==========

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics{{prior_omnuh2_tau_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on $A_{s}$. The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
\label{fig:prior_omeganuh2-50-as}
\end{center}
\end{figure}
%==========
%
%%==========
%\begin{figure}[htbp]
%\begin{center}
%\includegraphics{{prior_omnuh2_hubble_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
%\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the Hubble parameter $H_{0}$. The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
%\label{fig:prior_omeganuh2-50-h}
%\end{center}
%\end{figure}
%%==========
%
%
%%==========
%\begin{figure}[htbp]
%\begin{center}
%\includegraphics{{prior_omnuh2_omch2_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
%\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on $\Omega_{c} h^{2}$. The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
%\label{fig:prior_omeganuh2-50-omch2}
%\end{center}
%\end{figure}
%%==========
%
%%==========
%\begin{figure}[htbp]
%\begin{center}
%\includegraphics{{prior_omnuh2_re_optical_depth_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
%\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the optical depth $\tau$. The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
%\label{fig:prior_omeganuh2-50-tau}
%\end{center}
%\end{figure}
%%==========
%

@@

\reffig{prior_omeganuh2-h} shows the projected errors on $\sum m_\nu$ from CMB-S4 (+Planck+current BAO) as a function of external priors on $H_{0}, \Omega_{c}h^{2}, n_s$ and $A_s$. 
Without any external priors (far right in figure), upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 25 meV, corresponding to a 2-sigma detection even in the worst case scenario. Adding in external priors helps this significantly though, as discussed, for example, in \cite{di-valentino:2015,2013arXiv1309.5383A,pan:2015a,allison:2015}. 
There, the added prior was ``DESI BAO'', representing a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by the planned experiment DESI \cite{levi:2013}. This is represented by the blue band in our figures. In addition to showing the implications of introducing this fixed external constraint, \reffig{prior_omeganuh2-h} generalizes the extent to which external information will improve the neutrino mass constraints. For example, in \reffig{prior_omeganuh2-h} the red solid curve shows the improvement as additional information on the Hubble constant is applied. We can see, for example that, imposing {\it external} information at the level of $\sigma_{H_0}/H_0=0.5\%$ would improve the neutrino mass constraint to 17 meV. Similarly, in \reffig{prior_omeganuh2-omch2} adding to CMB S4 + BAO15 an external constraint on the dark matter density equal to its internal constraint of $0.4\%$ improves the projected neutrino mass constraint to 20 meV. It is not surprising that this is close to the constraint obtained with DESI because the BAO distances in this model depends most sensitively on $\Omega_ch^2$ and $H_{0}$.

The physical reason underlying this improvement is simple: lensing constrains $f_\nu$, the ratio of the neutrino density to the matter density. In order to extract the neutrino density (and therefore the neutrino mass), we need to know the matter density. A prior on $H_{0}$ helps for a related, but subtle reason. 
The position of the CMB peaks, well constrained by measurement, in a flat Universe depends on the combination $\Omega_ch^{2.93}$ \cite{planck-collaboration:2014}.
Therefore, an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$, leading back to tighter neutrino mass constraints.
As found in \cite{2013arXiv1309.5383A,pan:2015a,allison:2015}, the projected error falls below 20 meV with these priors. 
The left-hand side of \reffig{prior_omeganuh2-h} and \reffig{prior_omeganuh2-omch2} shows that even tighter external constraints on either $H_0$ or $\Omega_ch^2$ would significantly improve the neutrino mass constraints.

@@

Because of the presence of foregrounds, it may be difficult to get the very low-$\ell$ modes from the ground with CMB-S4, so in Fig.\ref{fig:prior_omeganuh2-50-as}-\ref{fig:prior_omeganuh2-50-tau} we limit the CMB S4 modes to $\ell>\ell_{\rm min}=50$. For the large scale modes $\ell<\ell_{\rm min}=50$ we use Planck at the level of accuracy forecasted by the Planck Blue Book.  
Together with the importance of $H_{0}$ and $\Omega_ch^2$ previously observed, in this case external priors on $A_{s}$ and $\tau$ are found to be helpful (see \reffig{prior_omeganuh2-50-ns} and \reffig{prior_omeganuh2-50-as}).

@@

The sensitivity to relativistic degrees of freedom comes from the effect of extra species on the damping tail of the CMB anisotropies \cite{2013PhRvD..87h3008H}, both in temperature and polarization, so parameters that also strongly affect this part of the spectrum, like the slope $n_s$ and the baryon density $\Omega_bh^2$, are the most relevant to improve the $N_{\rm eff}$ constraint. As such, \reffig{prior_neff-ombh2} and \reffig{prior_neff-ns} show that obtaining external priors on either of these would reduce the errors on $N_{\rm eff}$ to get close to a 4-sigma detection of the partial decoupling prediction.

@@

Toward that goal, we investigated here how these constraints depend on external priors:
\begin{description}
\item[$\sum m_\nu$] To improve the CMB constraint on the sum of neutrino masses by a factor of two, low-redshift measurements will be needed. Indeed an external prior on $\Omega_{c}h^{2}$ and $H_{0}$ at the level attained by CMB-S4 (sub-percent) can reduce the 1-$\sigma$ uncertainty by almost a factor of two.
Furthermore, if the final design of CMB S4 will not include measurements on large scales ($\ell<50$), external constraints on the primordial amplitude $A_{s}$ will also be useful. 
\item[$N_{\rm eff}$] The constraint on the relativistic degrees of freedom will benefit from external priors on parameters that affect the shape of the damping tail, such as the spectral index $n_s$ and the baryon energy density $\Omega_b h^2$.
\item[$w$] In this case the improvement will mainly come from external priors on $H_{0}$. For example a $1\%$ prior from an external experiment will reduce the error on $w$ by a factor of two.\end{description}

Much of the improvement that would come on neutrino masses and $w$ from $H_0$ or $\Omega_ch^2$ priors is anticipated to be provided by distance measurements made by DESI. But, additional $H_0$ measurements beyond DESI would help, as evidenced by the sharply falling curves in the left-most regions of \reffig{prior_omeganuh2-h}, \reffig{prior_omeganuh2-50-h} and \reffig{ellipse_w_H0}. Indeed, perhaps the most tantalizing finding is that future surveys beyond DESI and LSST \cite{lsst-dark-energy-science-collaboration:2012} that would tighten the constraints on, say, the baryon density and $n_s$ could indirectly lead to constraints in $N_{\rm eff}$ much less than a percent. Given the power of these constraints to probe physics beyond the Standard Model, such surveys seem particularly interesting to pursue.

>>>>

We also include information from current Baryon Acoustic Oscillation experiments and projected constraints from low-$l$ Planck polarization, as this is or will shortly be available.
For current BAO data (labeled BAO15), we follow \cite{allison:2015} and include 6dFGRS \cite{beutler:2011}, SDSS MGS \cite{ross:2015} together with LOW-Z and C-MASS \cite{anderson:2014}.
The noise assumed for Planck corresponds to the ``Planck-pol'' specifications of \cite{allison:2015}, where noise levels were approximated by scaling the current sensitivities according to the Planck Blue Book.

@@

%==========
\begin{figure}[htbp]
\begin{center}
\includegraphics{{prior_omnuh2_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
\caption{Projected constraints on the sum of the neutrino masses (absolute on the left and relative on the right) from CMB-S4 and current BAO measurements as a function of priors on the other cosmological parameters. The {\it internal} (marginalized constraints from CMB-S4 and current BAO) constraints on each parameter are given as $\sigma_{\rm pipeline}$. The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
\label{fig:prior_omeganuh2-h}
\end{center}
\end{figure}
%==========

%%==========
%\begin{figure}[htbp]
%\begin{center}
%\includegraphics{{prior_omnuh2_notau_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
%\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on $\Omega_{c} h^{2}$. The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
%\label{fig:prior_omeganuh2-omch2}
%\end{center}
%\end{figure}
%%==========

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics{{prior_omnuh2_lmin=50_lmax=4499_ndet=1000000_fsky=0.75_test}.pdf}
\caption{Same as Fig.~1, but with $l<50$ modes excluded from CMB-S4, instead coming from Planck. Note the importance of obtaining information on the optical depth $\tau$ if the low-$l$ modes are not measured accurately.}
\label{fig:prior-50}
\end{center}
\end{figure}
%==========
%
%%==========
%\begin{figure}[htbp]
%\begin{center}
%\includegraphics{{prior_omnuh2_hubble_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
%\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the Hubble parameter $H_{0}$. The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
%\label{fig:prior_omeganuh2-50-h}
%\end{center}
%\end{figure}
%%==========
%
%
%%==========
%\begin{figure}[htbp]
%\begin{center}
%\includegraphics{{prior_omnuh2_omch2_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
%\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on $\Omega_{c} h^{2}$. The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
%\label{fig:prior_omeganuh2-50-omch2}
%\end{center}
%\end{figure}
%%==========
%
%%==========
%\begin{figure}[htbp]
%\begin{center}
%\includegraphics{{prior_omnuh2_re_optical_depth_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
%\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the optical depth $\tau$. The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
%\label{fig:prior_omeganuh2-50-tau}
%\end{center}
%\end{figure}
%%==========
%

@@

\reffig{prior_omeganuh2-h} shows the projected errors on $\sum m_\nu$ from CMB-S4 (+Planck+current BAO) as a function of external priors on $H_{0}, \Omega_{c}h^{2}, n_s$ and $A_s$. 
Without any external priors (far right in figure), upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 25 meV, corresponding to a 2-sigma detection even in the worst case scenario. Adding in external priors helps this significantly though, as discussed, for example, in \cite{di-valentino:2015,2013arXiv1309.5383A,pan:2015a,allison:2015}. 
There, the added prior was ``DESI BAO'', representing a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by the planned experiment DESI \cite{levi:2013}. This is represented by the blue band in our figures. In addition to showing the implications of introducing this fixed external constraint, \reffig{prior_omeganuh2-h} generalizes the extent to which external information will improve the neutrino mass constraints. For example, in \reffig{prior_omeganuh2-h} the red solid curve shows the improvement as additional information on the Hubble constant is applied. We can see, for example that, imposing {\it external} information at the level of $\sigma_{H_0}/H_0=0.5\%$ would improve the neutrino mass constraint to 17 meV. It is not surprising that this is close to the constraint obtained with DESI because the BAO distances in this model depends most sensitively on $\Omega_ch^2$ and $H_{0}$.
Similarly, adding to CMB S4 + BAO15 an external constraint on the dark matter density equal to its internal constraint of $0.4\%$ improves the projected neutrino mass constraint to 20 meV. These are multi-dimensional examples of the familiar fact that adding two sets of equal constraints (in this case internal to S4+BAO15 and external) reduces the overall error by a factor of $\sqrt{2}$.

The physical reason underlying the improvement from $\Omega_ch^2$ is simple: lensing constrains $f_\nu$, the ratio of the neutrino density to the matter density. In order to extract the neutrino density (and therefore the neutrino mass), we need to know the matter density. A prior on $H_{0}$ helps for a related, but subtle reason. 
The position of the CMB peaks, well constrained by measurement, in a flat Universe depends on the combination $\Omega_ch^{2.93}$ \cite{planck-collaboration:2014}.
Therefore, an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$, leading back to tighter neutrino mass constraints.
As found in \cite{2013arXiv1309.5383A,pan:2015a,allison:2015}, the projected error falls below 20 meV with these priors. 
The left-hand side of \reffig{prior_omeganuh2-h} shows that even tighter external constraints on either $H_0$ or $\Omega_ch^2$ would significantly improve the neutrino mass constraints.

@@

Because of the presence of foregrounds, it may be difficult to get the very low-$\ell$ modes from the ground with CMB-S4, so in \reffig{prior-50}, we limit the CMB S4 modes to $\ell>\ell_{\rm min}=50$. For the large scale modes $\ell<\ell_{\rm min}=50$ we use Planck at the level of accuracy forecasted by the Planck Blue Book.  
Together with the importance of $H_{0}$ and $\Omega_ch^2$ previously observed, in this case external priors on $A_{s}$ and $\tau$ are found to be important. % (see \reffig{prior_omeganuh2-50-ns} and \reffig{prior_omeganuh2-50-as}).

@@

The sensitivity to relativistic degrees of freedom comes from the effect of extra species on the damping tail of the CMB anisotropies \cite{2013PhRvD..87h3008H}, both in temperature and polarization, so parameters that also strongly affect this part of the spectrum, like the slope $n_s$ and the baryon density $\Omega_bh^2$, are the most relevant to improve the $N_{\rm eff}$ constraint. As such, \reffig{prior_neff-ombh2} and \reffig{prior_neff-ns} show that obtaining external priors on either of these would reduce the errors on $N_{\rm eff}$ to get close to a 4-sigma detection of the partial decoupling prediction. Besides futuristic high redshift 21 cm ideas, the best bet to improve the constraints on $n_s$ are probes that push to redshift $z>1$ such as the Lyman alpha forest. There are many more linear Fourier modes available at high $z$ that can be used to obtain the larger level arm from large to small scales that is essential to extract high precision measurements of $n_s$.

@@

Toward that goal, we investigated here how these constraints depend on external priors:
\begin{description}
\item[$\sum m_\nu$] To improve the CMB constraint on the sum of neutrino masses by a factor of two, low-redshift measurements will be needed. Indeed an external prior on $\Omega_{c}h^{2}$ and $H_{0}$ at the level attained by CMB-S4 (sub-percent) can reduce the 1-$\sigma$ uncertainty by almost a factor of two.
Furthermore, if the final design of CMB S4 does not include measurements on large scales ($\ell<50$), external constraints from 21 cm experiments or all-sky CMB experiments on the optical depth $\tau$ will become very important.% powerful.
\item[$N_{\rm eff}$] The constraint on the relativistic degrees of freedom will benefit from external priors on parameters that affect the shape of the damping tail, such as the spectral index $n_s$. These are most likely to come from surveys that measure a large fraction of the $z>1$ linear Fourier modes.
\item[$w$] In this case the improvement will mainly come from external priors on $H_{0}$. For example a $1\%$ prior from an external experiment will reduce the error on $w$ by a factor of two.\end{description}

Much of the improvement that would come on neutrino masses and $w$ from $H_0$ or $\Omega_ch^2$ priors is anticipated to be provided by distance measurements made by DESI. But, additional $H_0$ measurements beyond DESI would help, as evidenced by the sharply falling curves in the left-most regions of \reffig{prior_omeganuh2-h}, \reffig{prior_omeganuh2-50-h} and \reffig{ellipse_w_H0}. Indeed, perhaps the most tantalizing finding is that future surveys beyond DESI and LSST \cite{lsst-dark-energy-science-collaboration:2012} that would tighten the constraints on, say, $n_s$ could indirectly lead to constraints in $N_{\rm eff}$ much less than a percent. Given the power of these constraints to probe physics beyond the Standard Model, such surveys seem particularly interesting to pursue.
